Staged Abstract Interpreters

### 2018-11-11 paper/22_bg_abs.tex
fixpoint iter

@@

\paragraph{Fixpoint Iteration.}
The interpreter may not terminate
co-inductive caching/truncated depth-first evaluation \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800, Rosendahl:AbsIntPL}

\begin{lstlisting}
object AbsInterp extends Abstract { // continued
  case class CacheFix(evev: EvalFun => EvalFun) {
    var in = Map[Config, Ans](); var out = Map[Config, Ans]()
    def cached_ev(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = {
      val cfg: Config = (e, $\rho$, $\sigma$)
      if (out.contains(cfg)) out(cfg)
      else {
        val ans0: Ans = in.getOrElse(cfg, Lattice[(Value, Store)].bot)
        out = out + (cfg -> ans0)
        val ans1: Ans = evev(cached_ev)(e, $\rho$, $\sigma$)
        out = out + (cfg -> Lattice[(Value, Store)].$\sqcup$(ans0, ans1))
        ans1
      }
    }
    def iter(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = {
      in = out; out = Map[Config, Ans](); cached_ev(e, $\rho$, $\sigma$)
      if (in == out) out((e, $\rho$, $\sigma$)) else iter(e, $\rho$, $\sigma$)
    }
  }
  override def eval_top(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    CacheFix(eval).iter(e, $\rho$, $\sigma$)
}
\end{lstlisting}

other big-step abstract interpreter, 
big-step abstract interpreter using delimited control operators \cite{Wei:2018:RAA:3243631.3236800},
monadic big-step abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17},
arrow-based abstract interpreter \cite{Keidel:2018:CSP:3243631.3236767}.

>>>>

\todo{maybe put to related work} other big-step abstract interpreter, 
big-step abstract interpreter using delimited control operators \cite{Wei:2018:RAA:3243631.3236800},
monadic big-step abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17},
arrow-based abstract interpreter \cite{Keidel:2018:CSP:3243631.3236767}.

@@

\paragraph{Fixpoint Iteration.}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also called co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL} and widely used in different analysis.

The idea is to set up two caches called @in@ and @out@, which are both mapping from the arguments of interpreter 
(@Config@) to the returned value of interpreter (@Ans@). The @in@ contains what we already computed from the last
iteration, the @out@ cache is what we computed after this iteration joined with the previous one.
When starting a new iteration, the @in@ is set to be the result of last iteration, i.e., @out@; the @out@ is set
to be empty. In the iteration, we first ask whether @out@ contains what we want, if yes then it is returned;
otherwise, we retrieve what we have from @in@, compute the result for this time, and put the joined result back to @out@.
Note that we also instrument the recursive call by putting @cached_ev@ to @evev@.
After one iteration, if @in == out@, then there is no more information can be gained, thus the iteration should 
end and we have reached the fixed point.

\begin{lstlisting}
case class CacheFix(evev: EvalFun => EvalFun) {
  var in = Map[Config, Ans](); var out = Map[Config, Ans]()
  def cached_ev(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = {
    val cfg: Config = (e, $\rho$, $\sigma$)
    if (out.contains(cfg)) out(cfg)
    else {
      val ans0 = in.getOrElse(cfg, Lattice[(Value, Store)].bot)
      out = out + (cfg -> ans0)
      val ans1 = evev(cached_ev)(e, $\rho$, $\sigma$)
      out = out + (cfg -> (ans0 $\sqcup$ ans1));  ans1
    }
  }
  def iter(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = {
    in = out; out = Map[Config, Ans](); cached_ev(e, $\rho$, $\sigma$)
    if (in == out) out((e, $\rho$, $\sigma$)) else iter(e, $\rho$, $\sigma$)
  }
}
override def eval_top(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
  CacheFix(eval).iter(e, $\rho$, $\sigma$)
\end{lstlisting}

Finally, we override the definition of @eval_top@ by instantiating @CacheFix@ with @eval@ and start 
the first iteration.

### 2018-11-11 paper/23_bg_stage.tex
fixpoint iter

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. \todo{NumV}
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, 
however, what we desire is the a \textit{compiled} closure, i.e., the syntactic term of the body expression should be
eliminated and specialized.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which both will be
exhibited in the generated code.

>>>>

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. \todo{NumV}
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, 
however, what we desire is the a \textit{compiled} closure, i.e., the syntactic term of the body expression should be
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which both will be
exhibited in the generated code.
